Supplementary Information

### 2021-01-07 doc/manuscript/MEE_manuscript.Rmd
worked on discussion

3. Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$. The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested). After accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (#%).

@@

**Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size and year on annual tree growth.** 
This allows us to ask:
*(1) What are the most important climate drivers of annual growth, and over which time windows?*
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
(4) How do climate drivers and $DBH$ additively and interactively shape growth? 
(5) How have growth rates changed through time?

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Jan6)
n_tested=37 #same for all metrics
n_best_model_RW=30
n_best_model_BAI=31
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,

@@

**Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental drivers on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) and with our current understanding of forest dynamics.** 
Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
Additive effects of water and temperature variables, usually over different time windows, occurred for most species, and nonlinear responses were common.
Interactions between climate variables and $DBH$ not unusual (Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`), with larger trees tending to be more drought sensitive (REFS).
Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ and was dependent upon species ecology in interaction with stand history and global change drivers (Figs. `r size`, `r year_fig`).
It is notable that growth rates of most species declined...
While none of these findings are biologically surprising, the simultaneous modeling of growth responses to climate, $DBH$, and year offers a fuller picture of the drivers of tree growth in an era of global change.

## Climate sensitivity {-#Climate}
*(corroboration of insight from conventional methods)*

**Our results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).**
An advantage to our approach is that that the use of *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.
Note that the use of *climwin* itself is analytically convenient, but does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* could also be used on residual chronologies (**AND WE DID THIS--  ADD SI FIGURE AS EXAMPLE?**). 
Where our approach differs fundamentally from traditional methods is in analyzing climate sensitivity of the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged. 
The result is that our approach yields climate correlations with greater variance but lower standard error of slope estimates (Figs. `r all_traditional_comparisons`).
Differences in slope between the two approaches are expected giving the methodological differences (Appendix `r traditional_comparison`), and were rarely statistically significant. 
(*add some discussion on why we expect different slopes?*)
Another way in which the current analysis differed from conventional methods is that we pooled species when determining the top climate drivers (*climwin* step; Fig. `r schematic`). 
This decision was motivated by the expectation that differences in optimal climate windows across species in one site (e.g., Fig. `r SCBI_example`) would be minimal compared to cross-site differences, and for the practical consideration that determining optimal climate windows for each species would be computationally intensive and challenging to present.
However, species-specific climate sensitivity estimates could be optimized by fitting *climwin* individually be species.
While our method sometimes may sometimes pick out different climate variables and time windows than would be identified via conventional methods, as a whole the identified drivers and directions of response are consistent with conventional methods (Table `r qualitative_results_comparison`).

**While yielding broadly similar results to conventional analyses, our analysis yielded new insights that would not have been found with conventional methods.**

1. additive effects are prevalent, and time windows often differ (trees tend to be sensitive to water over longer time scales (makes sense-- lags caused by soil moisture storage)

2. nonlinear effects are prevalent

3. DBH interactions --also potentially attributable to year-- are common

## Size trends {-#Size}

**The diversity of growth trends in relation to $DBH$ observed here (Fig. `r size`) is largely attributable to species ecology and stand history (Fig. 4).**
On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline. 
The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern. *).
On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]. 
While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al. 1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
--  (*cite Sheil et al. 2017 [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
-- *from PB: I think the one thing that is very hard to take into account (and which you noted in bold at the start of the paragraph) is stand history. I worry about looking for/finding general patterns just on tree size, as that ignores the historical dynamics that might influence these patterns.*

@@

Growth rates are changing directionally for most species/ sites.

>>>>

3. Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$. The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested). After accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (90%).

@@

**Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size and year on annual tree growth.** 
This approach allows us to ask:
(1) What are the most important climate drivers (and their time windows), and what is the shape of the relationship between annual growth and these drivers?
(2) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
(3) How do climate drivers and $DBH$ additively and interactively shape growth? 
(4) How have growth rates changed through time?

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Jan6)
n_tested=37 #same for all metrics
n_best_model_RW=30
n_best_model_BAI=31
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.3%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,

@@

Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by this approach.

Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
*(Discuss how this fits with regional to global-scale studies)*

While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), the approach differs in some substantive ways.
First, in determining the most important climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.

Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were *common/ prevalent*. 
*(DISCUSS)*

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%; Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

## Variation with $DBH$  {-#Variation}

**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$, with the observed diversity of growth trends in relation to $DBH$ (Fig. `r size`) largely attributable to species ecology and stand history.**
On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline. 
The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern. *).
On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]. 
While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al. 1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
--  (*cite Sheil et al. 2017 [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
-- *from PB: I think the one thing that is very hard to take into account (and which you noted in bold at the start of the paragraph) is stand history. I worry about looking for/finding general patterns just on tree size, as that ignores the historical dynamics that might influence these patterns.*

@@

It is notable that growth rates are changing directionally for most species/ sites.

### 2021-01-07 doc/manuscript/MEE_manuscript_SI.Rmd
Update MEE_manuscript_SI.Rmd
#35
worked on discussion

Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below). 
Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.
*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.

@@

We defined chronology start dates according to the subsample signal strength (SSS), using a cutoff of SSS = 0.80 (or 80% of the population signal).
Thus, for this analysis only, we defined chronology start dates as the year the SSS exceeded 0.80 or two years after the start of the climate record, whichever came later. 
SSS exceeded 0.80 well before the start of the 1901 start of climate records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU, SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
We note that these start date criteria differ from those used in the main analysis (Table `r sampling_details_table`), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
End dates were defined as the last full year prior to sampling (Table `r sampling_details_table`).

@@

>>>>

Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* could also be used on residual chronologies (**AND WE DID THIS--  ADD SI FIGURE AS EXAMPLE?**). 
Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.
*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.

@@

We defined chronology start dates according to the subsample signal strength (SSS), using a cutoff of SSS = 0.80 (or 80% of the population signal).
Thus, for this analysis only, we defined chronology start dates as the year the SSS exceeded 0.80 or two years after the start of the climate record, whichever came later. 
SSS exceeded 0.80 well before the start of the 1901 start of climate records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU, SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
We note that these start date criteria differ from those used in the main analysis (Table `r sampling_details_table`), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
End dates were defined as the last full year prior to sampling (Table `r sampling_details_table`), or 2007 for CB.

@@

*The result is that our approach yields climate correlations with greater variance but lower standard error of slope estimates (Figs. `r all_traditional_comparisons`).*
